---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```




# LineageHomology

  * [Installation](#installation)
  * [Example](#example)
  * [More advanced example and illustrations of some of the plotting methods](#for-a-more-advanced-example-and-illustrations-of-some-of-the-plotting-methods-see)

<!-- badges: start -->
<!-- badges: end -->

LineageHomology provides a set of functions that analyzes the outputs from an ancestral state reconstruction. Following  du Plessis et al. (2021). ( 10.1126/science.abf2946 )to estimate if taxa have inhereted traits by homology 
or if their character state represents a novelty wrt. the other taxa in the tree

## Installation

You can install the latest version of LineageHomology this repository using 

``` r
library(devtools)
devtools::install_github("magnusnosnes/LineageHomology")
```

## Example

This example shows how an reconstructed ancestral state history can be used to estimate local transmission lineages.


```{r,fig.width=10,fig.height=7}
library(LineageHomology)
## basic example code

#Loading other packages for simulating data. 
library(ape)
library(phytools)
library(phangorn)
library(BactDating)

#Simulate data from Norway and rest of the world
set.seed(400)
tree_test = simdatedtree(nsam=10, dateroot=2000)
tree_test = ladderize(tree_test)
Q=matrix(c(0.5,0.5,0.5,0.5), nrow=2,ncol=2, byrow=F)
colnames(Q)=c("Norway","RoW")
loc = c("Norway", "Norway","Norway","RoW", "RoW", "Norway", "Norway", "RoW", "RoW", "RoW")
names(loc) = tree_test$tip.label

#Reconstruct ancestral states using ace. 
fit1 = ace(x=loc, phy= tree_test, type="discrete", mod="ARD")
plot.phylo(tree_test,lwd=2,label.offset = 0.15, mar=c(0.2,0.2,0.2,0.2))
axisPhylo(root.time=2000, backward=F)
nodelabels(pie=fit1$lik.anc,cex=0.7,piecol=c("Red","Blue"))
tips = to.matrix(loc,seq=c("Norway", "RoW"))
tiplabels(pie=tips, cex=0.7,piecol=c("Red","Blue"))

LineageHomology(tree_test, ace_nodes=fit1$lik.anc,
                        ace_tips = to.matrix(loc, seq=c("Norway", "RoW")), start_time=2000)



```

#### For a more advanced example and illustrations of some of the plotting methods see

[Plotting tutorial link](https://github.com/magnusnosnes/LineageHomology/blob/master/Examples_and_plotting_methods/Simple_example/Basic_plotting.md)

