---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```




# LineageHomology

  * [Installation](#installation)
  * [Introduction](#introduction)
  * [Tutorial and gallery of plotting methods](#tutorial)

<!-- badges: start -->
<!-- badges: end -->

LineageHomology provides a set of functions that analyses the outputs from ancestral state reconstructions. LineageHomology takes the output of an ancestral state reconstruction method with included state probabilities at each node and defines transmission lineages (TLs) and singletons based on this. A TL is defined as a connected group of tips where state transitions between ancestral and descendant nodes have a probability lower than 50 percent. Singletons are defined as tips that are not connected to any other tips in this way. The method is analogous to that introduced by du Plessis et al. (2021) (DOI: 10.1126/science.abf2946). LineageHomology also provides descriptions of the sizes of TLs, the number of singletons, functions to derive estimates of importation and local transmission based on this and other useful summaries. The package also includes a variety of functions to plot the results. 

## Installation

You can install the latest version of LineageHomology from this repository by using devtools: 

```{r, eval=FALSE,echo=T}
library(devtools)
devtools::install_github("magnusnosnes/LineageHomology")
```

## Introduction

This introduction provides a simple example of applying ancestral state reconstruction to simulated geographical data for two locations and using LineageHomology to analyse the outputs. Here, LineageHomology provides descriptions of transmission lineages other plots the sizes and estimated time of the most recent common ancestor for each TL. 

First, we simulate data and estimate the ancestral geographical states:

```{r,fig.width=10,fig.height=7}
library(LineageHomology)
#Loading required packages.
library(ape)
#Loading other packages for simulating data. 
library(BactDating)

#Simulate data from Norway and rest of the world
set.seed(400)
tree_test = simdatedtree(nsam=10, dateroot=2000)
tree_test = ape::ladderize(tree_test)
Q=matrix(c(0.5,0.5,0.5,0.5), nrow=2,ncol=2, byrow=F)
colnames(Q)=c("Norway","RoW")
loc = c("Norway", "Norway","Norway","RoW", "RoW", "Norway", "Norway", "RoW", "RoW", "RoW")
names(loc) = tree_test$tip.label

#Reconstruct ancestral states using ace. 
fit1 = ace(x=loc, phy= tree_test, type="discrete", mod="ARD")
plot.phylo(tree_test,edge.width = 2,label.offset = 0.15, mar=c(0.2,0.2,0.2,0.2))
axisPhylo(root.time=2000, backward=F,lwd=2)
nodelabels(pie=fit1$lik.anc,cex=0.7,piecol=c("Red","Blue"))
tips = to.matrix(loc,seq=c("Norway", "RoW"))
tiplabels(pie=tips, cex=0.7,piecol=c("Red","Blue"))

```
The tree shows the reconstructed states using the ace function in the ape package. Each node is coloured according to the probability of the location. Here red represents Norway, and blue represents the rest of the world (RoW). 

Next we run lineageHomology on the output from ace. 
```{r,fig.width=10,fig.height=7}
Return = LineageHomology(tree_test, ace_nodes=fit1$lik.anc,
                        ace_tips = to.matrix(loc, seq=c("Norway", "RoW")), start_time=2000)
Return
```

In this example LineageHomology returned four transmission lineages. The taxa names of the tips included in each lineage is printed above under $Taxa_names. 
The size distributions can be visualised by, e.g. using a treemap plot:

```{r,fig.width=10,fig.height=7}
LineageHomology::treemap_lineagehomology(Return)
```

The figure shows squares with areas representing the transmission lineages' sizes. The text inside the squares gives the estimated time of the MRCA and the number of tips in the transmission lineages.


## Tutorial

See the full tutorial below for a full introduction to the package and more methods for visualizing the results.
[Tutorial and plotting methods](https://github.com/magnusnosnes/LineageHomology/blob/master/Examples_and_plotting_methods/Simple_example/Basic_plotting.md)
