---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%", warning=FALSE, message=F
)
```


## Using LineageHomology for a given phylogeny and with tips observed in geographical locations.

#### Table of contents 
* [Estimating local transmission linages for phylogeographic data with two categories.](#estimating-local-transmission-linages-for-phylogeographic-data-with-two-categories)
* [Simulate tip data with two locations](#simulate-phylogeographic-data-with-two-locations--norway-and-rest-of-the-world--row-)
* [Running LineageHomology](#running-lineagehomology)
* [Plotting transmission lineages](#plot-lineage-densities-over-time)
    - [Color the lineages after location](#we-can-color-the-groups-by-the-state-they-are-in--and-restrict-the-plotted-groups-to-sizes-larger-than-1-4--and-10)
* [Plot cumulative lineage size over time.](#plot-cumulative-lineage-size-over-time)
    - [Color the lineages after location](#again-we-can-add-color-to-the-groups-by-specifying-it)
* [Sample transmission lineages probabilistically](#probabilistic-counting-of-transmission-lineages)
* [Produce confidence interval on the largest transmission lineage](#estimate-size-of-largest-transmission-lineage)
* [Estimate importation and local transmission from phylogeographic results](#estimating-importation-and-local-transmission)



```{r,fig.width=10,fig.height=7}
#Load needed packages
library(LineageHomology)
library(ggplot2)
library(scales)
library(lubridate)

#Loading other packages for simulating data. 
library(ape)
library(phytools)
library(phangorn)
library(BactDating)

```

##### Simulate tip data with two different states.
For the purpose of this example, we let the geographical states represent Norway and the rest of the World (RoW). The package was originally developed geographical data with two locations.

```{r,fig.width=10,fig.height=15}
set.seed(400)
tree_test = simdatedtree(nsam=300, dateroot=2000) #300 taxa and date of the root at year 2000-
tree_test = ladderize(tree_test) #Reorder the tree to make it look nice
Q=matrix(c(0.5,0.5,0.5,0.5), nrow=2,ncol=2, byrow=F) #Set up a transition matrix for trait simulation
colnames(Q)=c("Norway","RoW") #From state 1 to state 2. 
rownames(Q)=c("Norway","RoW") #From state 1 to state 2. 

trait = phytools::sim.Mk(tree=tree_test,Q=Q,nsim=1) #Simulate the traits on the phylogeny

#Reconstruct ancestral states using ace. 
fit1 = ace(x=trait, phy= tree_test, type="discrete", mod="ARD") #Estimate the ancestral history

plot.phylo(tree_test,lwd=2,label.offset = 0.15, mar=c(0.2,0.2,0.2,0.2),cex=0.3) #Ploy phylogeny

axisPhylo(root.time=2000, backward=F) #Add time axis

nodelabels(pie=fit1$lik.anc,cex=0.2,piecol=c("Red","Blue"))
tips = to.matrix(trait,seq=c("Norway", "RoW"))
tiplabels(pie=tips, cex=0.2,piecol=c("Red","Blue"))

```
 
The pie charts on the nodes of the phylogeny represent the probability of the node being in the different geographical locations, where red represents Norway, and blue represents RoW. LineageHomology uses these probabilities to count connected groups of tips, where all the nodes on the paths between them have a probability that > 50% for the same geographical state. LineageHomology also returns the number of tips that are not connected to any other tips in this way.  Following du Plessis et al. (2021) (DOI: 10.1126/science.abf2946) we refer to the groups as transmission lineages (TLs) and isolated tips as singletons. LineageHomology also returns other useful summaries.

#### Running LineageHomology

Next we run LineageHomology. The inputs should take this format.
```{r,fig.width=10,fig.height=15}
tree_test
head(fit1$lik.anc)
head( to.matrix(trait, seq=c("Norway", "RoW")))

```

Note that the node probabilities of the tips must have row names that match the tips of the phylogeny.

```{r,fig.width=10,fig.height=15}
Result = LineageHomology(tree_test, ace_nodes=fit1$lik.anc,
                        ace_tips = to.matrix(trait, seq=c("Norway", "RoW")), start_time=2000)

```

LineageHomology outputs a list of lists. 

* Import_LocalTrans
contains an estimate of the number of taxa that are have changed state wrt. The state of the closest common ancestral node is shared with any other taxa. 
* Lineage_sizes contains the number of taxa that belong to the mapped state. This means that for the entire group, all the internal nodes will be mapped to the same state with a 50% or higher probability. 
*Taxa names contain sub-lists for each lineage, in the same order as the other lists. 
   + Each sublist contains the taxa names of all the taxa in the group. 
* MRCA's contains the time of the most recent common ancestor of the groups. 
* Lineage_state contains the state the groups are mapped to. The states are treated as numbers in the analyses, so the relative ordering has to be looked up.


```{r,fig.width=10,fig.height=7}
names(Result)
Result$Import_LocalTrans
Result$lineage_state
Result$Lineage_sizes
head(Result$Taxa_names)
Result$`MRCA's`
Result$lineage_state


```

#Make a treemap plot to get a quick overview of the lineages.

Treemap_lineagehomology uses the treemap package to visualize transmission lineages as rectangles with areas that reflect their relative sizes. Additionally, the text on each square gives the name of group number of the transmission lineage, the estimated time of the mrca (TMRCA) of the transmission lineage and the number of tips in the group (S)

```{r,fig.width=10,fig.height=7}
LineageHomology::treemap_lineagehomology(Result)
```

##### Plot lineage densities over time 
Before plotting, we process the results using the function lineage_info. lineage_info takes the results from LineageHomology and names and dates observed dates of the tips in the tree. The formats are shown below.  
```{r,fig.width=10,fig.height=7}
#Set up matrix with taxa info
name_date = data.frame(name = names(trait), dates= BactDating::leafDates(tree_test))
head(name_date)
Result_lineage_info = LineageHomology::lineage_info(Result,name_date)
```

Another plotting method is density plots of the transmission lineages. Ridgeplot_lineagedensities takes the input from the function lineage_info and produces density plots using for each TL over time. The density plots are made using the ggridges R-package. The TLs are sorted from largest to smallest and can be coloured according to the state that defines the TL. 

```{r,fig.width=10,fig.height=7}
LineageHomology::ridgeplot_lineagedensities(Result_lineage_info=Result_lineage_info,groups_larger_than = 1,datelims=c("2000-01-01","2025-01-01","3 year"),color_by_state = F)
```

##### We can colour the groups by the state they are in, and restrict the plotted groups to sizes larger than 1,4, and 10

```{r,fig.width=10,fig.height=7}

LineageHomology::ridgeplot_lineagedensities(Result_lineage_info=Result_lineage_info,groups_larger_than = 1,datelims=c("2000-01-01","2025-01-01","3 year"),color_by_state = T)

LineageHomology::ridgeplot_lineagedensities(Result_lineage_info=Result_lineage_info,groups_larger_than = 4,datelims=c("2000-01-01","2025-01-01","3 year"),color_by_state = T)

LineageHomology::ridgeplot_lineagedensities(Result_lineage_info=Result_lineage_info,groups_larger_than = 10,datelims=c("2000-01-01","2025-01-01","3 year"),color_by_state = T)


```
 
##### Plot cumulative lineage size over time. 

```{r,fig.width=10,fig.height=7}

LineageHomology::lineage_growth_cumulative(Result_lineage_info = Result_lineage_info, datelims=c("2000-01-01","2025-01-01","3 year"))

```
 
##### Again, we can add colour to the groups by specifying it.

```{r,fig.width=10,fig.height=7}
LineageHomology::lineage_growth_cumulative(Result_lineage_info = Result_lineage_info, datelims=c("2000-01-01","2025-01-01","3 year"),color_by_state = T)

```

### Probabilistic counting of transmission lineages

For all the figures shown above, the transmission lineage division was done by division of lineages at points where a transition is estimated with more than 50 percent probability. If one wishes to rather count transmission lineages by sampling from the observed probabilities, one can use the function LineageHomology_w_uncertainty.

```{r,fig.width=10,fig.height=7}
Result1 = LineageHomology_w_uncertainty_v2(tree_test, ace_nodes=fit1$lik.anc,
                        ace_tips = to.matrix(trait, seq=c("Norway", "RoW")), start_time=2000)
LineageHomology::treemap_lineagehomology(Result1)
Result2 =  LineageHomology_w_uncertainty_v2(tree_test, ace_nodes=fit1$lik.anc,
                        ace_tips = to.matrix(trait, seq=c("Norway", "RoW")), start_time=2000)
LineageHomology::treemap_lineagehomology(Result2)
```

As these figures show, the division of lineages will vary according to the sampled states of some of the nodes. This can be useful for producing confidence limits on, e.g. the size of the largest lineage.

### Estimate size of largest transmission lineage

We can use pbreplicate to counts transmission lineages probabilistically many times to obtain confidence intervals (or, in this case, a distribution) for the size of the largest transmission lineage.

```{r,fig.width=10,fig.height=7}
library(pbapply)
multi_counts = pbreplicate(
  1000,
  LineageHomology::LineageHomology_w_uncertainty_v2(
    tree=tree_test,
    ace_nodes=fit1$lik.anc,
    ace_tips = to.matrix(trait, seq=c("Norway", "RoW")),
    start_time = 2000
  )
)

largest_lineage = c()
for(i in 1:1000) {
  largest_lineage = c(largest_lineage, max(multi_counts[,i]$Lineage_sizes))
}

hist(largest_lineage, breaks=100, xlim=c(0,40), probability = F, main="Size of largest transmission lineage",xlab="Size", ylab="Probability")

```


#### Estimating importation and local transmission

Here we focus on Norway and run LineageHomology on the observed tips in Norway before the contribution of importation and local transmission events over time.

```{r,fig.width=10,fig.height=7}
library(pbapply)

Norwegian_tips = names(trait[trait=="Norway"])
head(Norwegian_tips)

multi_counts = pbreplicate(
  1000,
  LineageHomology::LineageHomology_w_uncertainty_v2(
    tree=tree_test,
    give_tips = Norwegian_tips,
    ace_nodes=fit1$lik.anc,
    ace_tips = to.matrix(trait, seq=c("Norway", "RoW")),
    start_time = 2000
  )
)


 #Count number of local transmission and importations in half-year (0.5) intervals.
result_import_local_transmission = import_local_transmission(tree = tree_test,LineageHomology_replicates = multi_counts, start_time = 2000,time_interval_size = 0.5)

```

We can summarise the number of tips estimated to be a result of importation events and local transmission events. 

```{r,fig.width=10,fig.height=7}
Summarize_import_local_transmission(LineageHomology_replicates = multi_counts)
```

Lastly, we can plot the relative contribution of importation and local transmission in time intervals spanning half a year. 

```{r,fig.width=10,fig.height=7}
plot_importation_local_transmission(tree = tree_test, result_import_local_transmission = result_import_local_transmission,start_time=2000, time_interval_size = 0.5, date_breaks = "2 years",importation_or_local = "both") #Use importation_or_local = "importation" or "local" if you do not have the grid package installed. 

```

#### Questions regarding the package can be directed to magnusnosnes@gmail.com
